Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations

### 2021-07-05 submission/manuscript.rmd

---
title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit"
abstract: |
 **Objective:** To compare different risk-based methods for predicting individualized treatment effects using simulations of the RCT setting. **Study Design and Setting:** We sampled patients using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic). In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in 4 quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive model selection approach based on Akaike’s Information Criterion. We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample size, treatment effect strength, and the PI's discriminative ability. **Results:** Models including a PI by treatment interaction performed better under most simulation settings. Flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model. The adaptive approach performed similarly to the best-performing method in most scenarios. **Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

>>>>

---
title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations"
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized treatment effects in RCT simulations. **Study Design and Setting:** We simulated data using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic). In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach using Akaike’s Information Criterion. We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect strength, and the PI's discriminative ability. **Results:** Models including a PI by treatment interaction performed better under most simulation settings. Flexible models required larger sample sizes and higher AUC of the PI to outperform the linear interaction model. The adaptive approach performed similarly to the best-performing method in most scenarios. **Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

### 2021-07-05 submission/manuscript.tex

\title{Models including an interaction of treatment with baseline risk
were better able to predict absolute benefit}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods for
predicting individualized treatment effects using simulations of the RCT
setting. \textbf{Study Design and Setting:} We sampled patients using
diverse assumptions for a baseline prognostic index of risk (PI) and the
shape of its intereraction with treatment (none, linear or quadratic).
In each sample we predicted absolute benefit using: models with the PI
and a constant relative treatment effect; models including an
interaction of treatment with the PI; stratification in 4 quarters of
the PI; nonlinear transformations of the PI (restricted cubic splines
with 3, 4 and 5 knots); an adaptive model selection approach based on
Akaike's Information Criterion. We evaluated predictive performance
using root mean squared error and measures of discrimination and
calibration for benefit. Starting from a base case scenario (sample size
4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample
size, treatment effect strength, and the PI's discriminative ability.
\textbf{Results:} Models including a PI by treatment interaction
performed better under most simulation settings. Flexible models with
restricted cubic splines required larger sample sizes and higher AUC of
the PI to outperform the linear model. The adaptive approach performed
similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Usually, a model with a linear interaction of the
PI with treatment is better able to predict absolute treatment benefit.
\end{abstract}

>>>>

\title{Models including an interaction of treatment with baseline risk
were better able to predict absolute benefit in simulations}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
performed similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Usually, a model with a linear interaction of the
PI with treatment is better able to predict absolute treatment benefit.
\end{abstract}
